A user-extensible and adaptable parser architecture

### 2008-07-29 doc/logparser-results.tex
Improvements copied from the thesis. Beginnings of a timing data section. Wrote a short piece about IBL. Some corrections from Carl. Tweaked graph sizes and spacing between floats to reduce page count.

When generating the timing data used in this section, \numberOFlogFILES{}
log files (totaling 10.08 GB, \numberOFlogLINEShuman{} log lines) were each
parsed 10 times, the results from the first first run discarded, and the
execution time for the remaining 9 runs averaged.  The results from the
first run were discarded because because the rule ordering will be
sub-optimal compared to subsequent runs --- XXX MAYBE I SHOULD NOT DISCARD
THE FIRST RESULT\@?  Saving results to the database was disabled for the
test runs, as that dominates the run time of the program, and the tests are
aimed at measuring the speed of the parser rather than the speed of the
database and the disks the database is stored on.

\subsubsection{Algorithmic complexity}

An important property of a parser is how execution time scales relative to
input size: does it scale linearly, polynomially, or exponentially?
\Graphref{parsing time vs file size vs number of lines graph} shows the
parsing time in seconds, file size in MB, and tens of thousands of log
lines for each of the \numberOFlogFILES{} log files.  All three lines run
roughly in parallel, giving the impression that the algorithm scales
linearly with input size.  This impression is borne out by
\graphref{parsing time vs file size vs number lines factor} which plots
both the ratio of file size vs parsing time, and the ratio of number of log
lines vs parsing time (higher is better in both cases); \tableref{parsing
time vs file size vs number lines factor table} shows the same ratios for
different groups of log files.  As the reader can see the ratios are quite
tightly banded, providing empirical evidence that the algorithm scales
linearly.  The ratio increases (i.e.\ improves) for log files 22 and 62--68
despite their larger than average size (shown in \graphref{parsing time vs
file size vs number of lines graph}); this strange behaviour where larger
log files are parsed more efficiently than smaller is explained fully in
\sectionref{Explaining the peaks in log file size}.

>>>>

When generating the timing data used in this section, \numberOFlogFILES{}
log files (totaling 10.08 GB, \numberOFlogLINEShuman{} log lines) were each
parsed 10 times, the results from the first first run discarded, and the
execution time for the remaining 9 runs averaged.  The results from the
first run were discarded because because the rule ordering will be
sub-optimal compared to subsequent runs --- XXX MAYBE I SHOULD NOT DISCARD
THE FIRST RESULT\@?  Saving results to the database was disabled for the
test runs, as that dominates the run time of the program, and the tests are
aimed at measuring the speed of the parser rather than the speed of the
database and the disks the database is stored on.  The computer used for
test runs was a Dell Optiplex 745, equipped with:

\begin{tabular}[]{ll}

CPU         & One dual core 2.40GHz Intel(R) Core(TM)2 CPU, \\
            & with 32KB L1 cache and 4MB L2 cache. \\
RAM         & 2GB 667 MHz DDR RAM. \\
Hard disk   & One XXX 232GB SATA disk. \\

\end{tabular}

In as far as possible the computer was dedicated to the task of gathering
statistics from test runs, and was not used for any other purpose while
test runs were ongoing.

\subsubsection{Architecture Scalability: Input Size}

An important property of a parser is how parsing time scales relative to
input size: does it scale linearly, polynomially, or exponentially?
\Graphref{parsing time vs file size vs number of lines graph} shows the
parsing time in seconds, file size in MB, and tens of thousands of log
lines for each of the \numberOFlogFILES{} log files.  All three lines run
roughly in parallel, giving the impression that the algorithm scales
linearly with input size.  This impression is borne out by
\graphref{parsing time vs file size vs number lines factor} which plots
both the ratio of file size vs parsing time, and the ratio of number of log
lines vs parsing time (higher is better in both cases); \tableref{parsing
time vs file size vs number lines factor table} shows the same ratios for
different groups of log files.  As the reader can see the ratios are quite
tightly banded, providing empirical evidence that the algorithm scales
linearly.  The ratio increases (i.e.\ improves) for log files 22 and 62--68
despite their larger than average size (shown in \graphref{parsing time vs
file size vs number of lines graph}); this strange behaviour where larger
log files are parsed more efficiently than smaller is explained fully in
\sectionref{Explaining the peaks in log file size}.

### 2008-07-29 papers/sgai-2008/paper.tex
Improvements copied from the thesis. Beginnings of a timing data section. Wrote a short piece about IBL. Some corrections from Carl. Tweaked graph sizes and spacing between floats to reduce page count.

@@

% \showgraph{filename}{caption}{label}
\newcommand{\showgraph}[3]{%
    \begin{figure}[hbt!]
        \caption{#2}\label{#3}
        \includegraphics{#1}
    \end{figure}
}

@@

@@

@@

The solution developed is conceptually simple: provide a few generic
functions (\textit{actions\/}), each capable of dealing with an entire
category of log lines (e.g.\ rejections), accompanied by a multitude of
precise patterns (\textit{rules\/}), each of which matches all log lines of
a specific type, and only that type (e.g.\ rejection by a specific DNSBL).
This architecture is ideally suited to parsing inputs where the input is
not fully understood or does not follow a fixed grammar: the architecture
warns about unparsed inputs and other errors, but continues on parsing as
best it can, allowing the developer of a new parser to decide which
deficiencies are most important and require attention first, rather than
being forced to fix the first error which arises.

@@

Each action contains the code required to deal with log lines from a
particular category, e.g.\ rejections.  The actions are parser-specific:
each parser will need to write the actions it requires from scratch, unless
it is extending an existing parser.  In the Postfix parser developed for
this project there are \numberOFactions{} actions and \numberOFrules{}
rules; the distribution of rules to actions is very skewed, as shown in
Fig.~\refwithpage{Distribution of rules per action}.
\showgraph{build/plot-action-distribution}{Distribution of rules per
action}{Distribution of rules per action} The distribution is skewed for
the same reason the architecture is so successful in enabling users to
parse new log lines: most new log lines can be parsed by a combination of a
new rule and an existing action.  Unsurprisingly the action with the
highest number of associated rules is \texttt{REJECTION}, the action which
handles rejections; it is followed by \texttt{SAVE\_BY\_QUEUEID}, the
action responsible for handling informative inputs: these are log lines
that supplement the data gathered from other log lines (such as
rejections), but that do not by themselves represent interesting events.
The third most common action is, perhaps surprisingly, \texttt{IGNORE}:
there are many log lines which are not of interest to the parser, e.g.\
warning messages about DNS mismatches.  The parser warns about every
unparsed log line, to alert the user that they need to alter or extend
their ruleset; the \texttt{IGNORE} action does nothing when executed,
allowing uninteresting log lines to be parsed without causing any effects.
The remaining actions have only one or two associated rules: in some cases
this is because there will only ever be one type of log line for that
action, e.g.\ all log lines showing that a remote client has connected are
matched by a single rule and handled by the \texttt{CONNECT} action; in
other cases the action is required to address a deficiency in the log
files, or a complication which arises during parsing.

@@

The ability to add special purpose actions to deal with difficulties and
new requirements which are discovered during parser development is one of
the strengths of this architecture: instead of writing one monolithic
function which must be modified to support new behaviour or resolve
challenges, with all the attendant risks of adversely affecting the
existing parser, when a new requirement arises an independent action can be
written to satisfy it.  In some cases the new action will require the
cooperation of other actions, e.g.\ to set or check a flag; there is a
possibility of introducing failure by modifying existing actions, but
modifications will be smaller and occur less frequently than with a
monolithic architecture, thus failures should be less likely, and easier to
test for and diagnose.  The parser can be written in an object oriented
style, allowing sub-classes to extend or override actions in addition to
adding new actions; because each action is an independent function, the
sub-class need only modify the action it is interested in, rather than
reproducing large chunks of functionality.

@@

Rules are responsible for categorising inputs, specifying both the data to
be extracted from each log line and the action that should be invoked.  The
Postfix log parser stores the rules in the same SQL database that the
results are stored in, removing any doubt as to which set of rules was used
to produce a set of results; other implementations are free to store the
rules in whatever fashion suits their needs.  Each rule defines several
attributes:

@@

    \item [program] The Postfix program whose log lines the rule should be
        applied to.  This avoids needlessly trying rules that will not
        match the log line, or worse, might match unintentionally.  Rules
        whose program is \texttt{*} will be tried against any log lines not
        parsed by program specific rules, allowing generic rules to be
        written.

@@

The rules used by the Postfix log file parser define several additional
attributes; each implementation is free to add any additional attributes it
requires.

Adding new rules is simple: simply define all the required attributes and
append the new rule to wherever the existing rules are stored.  The
software developed for this project includes a utility based on
SLCT~\cite{slct-paper} to aid in producing regexes to parse previously
unparsed inputs.

@@

Parsing efficiency is an obvious concern when the parser routinely needs to
deal with 75 MB log files containing 300,000 log lines (generated daily on
a mail server handling mail for approximately 700 users --- large scale
mail servers would produce much larger log files on a daily basis).  When
generating the data for the graphs included in this document,
\numberOFlogFILES{} log files (totaling 10.08 GB, \numberOFlogLINEShuman{}
log lines) were each parsed 10 times, the results from the first run
discarded, and the execution time for the remaining 9 runs averaged.

@@

An important property of a parser is how execution time scales relative to
input size: does it scale linearly, polynomially, or exponentially?
Figure~\refwithpage{Execution time, file size, and number of lines} shows
the execution time in seconds, file size in MB, and tens of thousands of
log lines per log file.  All three lines run roughly in parallel, giving
the impression that the algorithm scales linearly with input size.  This
impression is borne out by fig.~\refwithpage{execution time vs file size vs
number lines factor} which plots the ratios of file size vs execution time,
and number of log lines vs execution time (higher is better in both cases).
As the reader can see the ratios are quite tightly banded across the graph,
showing that the algorithm scales linearly: the much larger log files at
points 62--68 on the X axis in fig.~\refwithpage{Execution time, file size,
and number of lines} actually cause the ratio to increase (i.e.\ improve),
rather than decrease.  The strange behaviour where larger log files are
parsed more efficiently is explained in \sectionref{Why are there dips in
the graphs?}.  \showgraph{build/plot-normal-filesize-numlines}{Execution
time, file size, and number of lines}{Execution time, file size, and number
of lines} \showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of
file size and number of lines to execution time}{execution time vs file
size vs number lines factor}

@@

Figure~\refwithpage{Parsing time relative to shuffled} shows the percentage
increase of execution times for shuffled and reversed rules.  Overall this
optimisation provides a modest but worthwhile performance increase of over
10\%, for a small investment in time and programming.
\showgraph{build/plot-normal-and-reverse-relative-to-shuffle}{Parsing time
relative to shuffled}{Parsing time relative to shuffled}

@@

The dips at log files 22 and 62--68 in fig.~\refwithpage{Parsing time
relative to shuffled} correspond to peaks in log file size in
fig.~\refwithpage{Execution time, file size, and number of lines}, and
peaks in fig.~\refwithpage{execution time vs file size vs number lines
factor}, where a peak means that performance is better.  The explanation
for this took some time to arrive at, but it turns out to be reasonably
simple.  The large log files in question were caused by a mail forwarding
loop, where the distribution of log lines is quite different to normal,
resulting in different performance characteristics.  The vast majority of
log lines when a mail loop occurs are from Postfix components which have a
small number of rules associated with them, whereas in general the majority
of log lines are from the Postfix component (\daemon{smtpd}) with the
highest number of rules.  Furthermore in normal log files the majority of
log lines are distributed much more evenly across rules than the majority
of log lines during a mail loop.

These two characteristics combine to reduce the average number of rules
required to parse a log line when there is a mail loop, as shown by the
peaks in fig.~\refwithpage{execution time vs file size vs number lines
factor} and troughs in fig.~\refwithpage{Parsing time relative to
shuffled}.  When the rule ordering is reversed the majority of log lines
generated by a mail loop will be parsed using very few rules, because there
are very few rules for the Postfix components producing the log lines.
Without a mail loop the Postfix components producing the majority of log
lines have a large number of rules associated with them, and reverse
ordering deliberately starts with the least likely rules and moves towards
the most likely.  This leads to a noticeable drop in the average time
required to parse a log line during a mail loop, as shown in
fig.~\refwithpage{Parsing time relative to shuffled}.  The number of rules
which need to be consulted when the ordering is shuffled varies between the
optimal and the pessimal, with the performance varying proportionally.

>>>>

\setlength{\floatsep}{0pt}
\setlength{\intextsep}{0pt}
\setlength{\textfloatsep}{0pt}

@@

% \showgraph{filename}{caption}{label}
\newcommand{\showgraph}[3]{%
    \begin{figure}[hbt!]%
        \caption{#2}\label{#3}%
        \includegraphics{#1}%
    \end{figure}%
}

@@

\newcommand{\refwithlabel}[2]{%
    #1~\vref{#2}%
}
% table references, for consistent formatting.
\newcommand{\graphref}[1]{%
    \refwithlabel{fig.}{#1}%
}
\newcommand{\Graphref}[1]{%
    \refwithlabel{Figure}{#1}%
}

@@

\newcommand{\numberOFlogFILESall}[0]{%
    522%
}

@@

The solution developed is conceptually simple: provide a few generic
functions (\textit{actions\/}), each capable of dealing with an entire
category of log lines (e.g.\ rejections), accompanied by a multitude of
precise patterns (\textit{rules\/}), each of which matches all log lines of
a specific type, and only that type (e.g.\ rejection by a specific DNSBL).
It is an accepted standard to separate the parsing procedure from the
declarative grammar it operates with.  Part of the novelty here is that the
grammars (rules and actions) are themselves procedures.  This architecture
is ideally suited to parsing inputs where the input is not fully understood
or does not follow a traditional grammar: the architecture warns about
unparsed inputs and other errors, but continues on parsing as best it can,
allowing the developer of a new parser to decide which deficiencies are
most important and require attention first, rather than being forced to fix
the first error which arises.

@@

Each action contains the code required to deal with log lines from a
particular category, e.g.\ rejections.  The actions are parser-specific:
each parser author will need to write the actions required from scratch,
unless extending an existing parser.  In the Postfix parser developed for
this project there are \numberOFactions{} actions and \numberOFrules{}
rules; the distribution of rules to actions is very skewed, as shown in
Fig.~\refwithpage{Distribution of rules per action}.  The distribution is
skewed for the same reason the architecture is so successful in enabling
users to parse new log lines: most new log lines can be parsed by a
combination of a new rule and an existing action.
\showgraph{build/plot-action-distribution}{Distribution of rules per
action}{Distribution of rules per action} Unsurprisingly the action with
the highest number of associated rules is \texttt{REJECTION}, the action
which handles Postfix rejecting a mail; it is followed by
\texttt{SAVE\_BY\_QUEUEID}, the action responsible for handling informative
inputs: these are log lines that supplement the data gathered from other
log lines (such as rejections), but that do not by themselves represent
interesting events.  The third most common action is, perhaps surprisingly,
\texttt{IGNORE}: the \texttt{IGNORE} action does nothing when executed,
allowing uninteresting log lines to be parsed without causing any effects.
The parser warns about every unparsed log line, to alert the user that they
need to alter or extend their ruleset; the Postfix parser successfully
parses every input line in the \numberOFlogFILESall{} log files it is
currently being tested with.  The remaining actions have only one or two
associated rules: in some cases this is because there will only ever be one
type of log line for that action, e.g.\ all log lines showing that a remote
client has connected are matched by a single rule and handled by the
\texttt{CONNECT} action; in other cases the action is required to address a
deficiency in the log files, or a complication which arises during parsing.

@@

The ability to add special purpose actions to deal with difficulties and
new requirements which are discovered during parser development is one of
the strengths of this architecture: instead of writing one monolithic
function which must be modified to support new behaviour, with all the
attendant risks of adversely affecting the existing parser, when a new
requirement arises an independent action can be written to satisfy it.  In
some cases the new action will require the cooperation of other actions,
e.g.\ to set or check a flag; there is a possibility of introducing failure
by modifying existing actions, but modifications will be smaller and occur
less frequently than with a monolithic architecture, thus failures should
be less likely, and easier to test for and diagnose.  The parser can be
written in an object oriented style, allowing sub-classes to extend or
override actions in addition to adding new actions; because each action is
an independent function, the sub-class need only modify the action it is
interested in, rather than reproducing large chunks of functionality.

@@

Rules categorise inputs, specifying both the data to be extracted from each
log line and the action to invoke.  The Postfix log parser stores the rules
in the same SQL database that the results are stored in, removing any doubt
as to which set of rules was used to produce a set of results; other
implementations are free to store the rules in whatever fashion suits their
needs.

XXX ADD SOMETHING ABOUT IMPLEMENTATIONS BEING FREE TO ADD OR REMOVE
ATTRIBUTES\@.

@@

    \item [program] The Postfix program whose log lines the regex should be
        applied to.  This avoids needlessly trying regexes that will not
        match the log line, or worse, might match unintentionally.  Rules
        whose program is \texttt{*} will be tried against any log lines not
        parsed by program specific rules, allowing generic rules to be
        written.

@@

Adding new rules is simple: simply define all the required attributes and
append the new rule to wherever the existing rules are stored.  The
software developed for this project includes a utility based on
SLCT~\cite{slct-paper} to aid in producing regexes to parse previously
unparsed inputs.  Storing the rules separately allows other approaches to
rule management to be used, e.g.\ instead of manually adding new rules when
unparsed log lines are seen, machine learning techniques could be used to
automatically generate new rules, based on the existing rules and input
lines.  If this approach was taken the choice of machine learning technique
would be constrained by the size of typical data sets: the
\numberOFlogFILES{} log files used to generate statistics contain
\numberOFlogLINEShuman{} log lines.  Techniques which require the full data
set when training would be impractical; Instance Based Learning
(IBL)~\cite{instance-based-learning} techniques which automatically
determine which inputs from the training set are valuable and which can be
discarded would probably reduce the data required to a manageable size.
This avenue of research and development has not been pursued by the author,
but could easily be independently undertaken.

@@

Parsing efficiency is an obvious concern when the parser routinely needs to
parse very large log files.  The server which generated the log files used
in testing this parser accepts approximately 10,000 mails for 700 users a
day.  Median log file size is 50MB, containing 285,000 log lines --- large
scale mail servers would have much larger log files.  When generating the
timing data used in this section, \numberOFlogFILES{} log files (totaling
10.08 GB, \numberOFlogLINEShuman{} log lines) were each parsed 10 times,
the results from the first first run discarded, and the execution time for
the remaining 9 runs averaged.  Saving results to the database was disabled
for the test runs, as that dominates the run time of the program, and the
tests are aimed at measuring the speed of the parser rather than the speed
of the database and the disks the database is stored on.  The computer used
for test runs was a Dell Optiplex 745, equipped with:

\begin{tabular}[]{ll}

CPU         & One dual core 2.40GHz Intel(R) Core(TM)2 CPU, \\
            & with 32KB L1 cache and 4MB L2 cache.          \\
RAM         & 2GB 667 MHz DDR RAM\@.                        \\
Hard disk   & One XXX 232GB SATA disk.                      \\

\end{tabular}

In as far as possible the computer was dedicated to the task of gathering
statistics from test runs, and was not used for any other purpose while
test runs were ongoing.

XXX TIMING INFORMATION\@.
\newpage

@@

An important property of a parser is how parsing time scales relative to
input size: does it scale linearly, polynomially, or exponentially?
\Graphref{Parsing time, file size, and number of lines} shows the parsing
time in seconds, file size in MB, and tens of thousands of log lines for
each of the \numberOFlogFILES{} log files.  All three lines run roughly in
parallel, giving the impression that the algorithm scales linearly with
input size.  This impression is borne out by \graphref{parsing time vs file
size vs number lines factor} which plots both the ratio of file size vs
parsing time, and the ratio of number of log lines vs parsing time (higher
is better in both cases).  As the reader can see the ratios are quite
tightly banded across the graph, showing that the algorithm scales
linearly.  The ratio increases (i.e.\ improves) for log files 22 and 62--68
despite their larger than average size (shown in \graphref{Parsing time,
file size, and number of lines}); this strange behaviour where larger log
files are parsed more efficiently is explained in \sectionref{Why are there
dips in the graphs?}.
\showgraph{build/plot-normal-filesize-numlines}{Parsing time, file size,
and number of lines}{Parsing time, file size, and number of lines}
\showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file size
and number of lines to parsing time}{parsing time vs file size vs number
lines factor}

@@

Figure~\refwithpage{Parsing time relative to shuffled} shows the percentage
increase of parsing times for shuffled and reversed rules.  Overall this
optimisation provides a modest but worthwhile performance increase of over
10\%, for a small investment in time and programming.
\showgraph{build/plot-normal-and-reverse-relative-to-shuffle}{Parsing time
relative to shuffled}{Parsing time relative to shuffled}

@@

The dips at log files 22 and 62--68 in fig.~\refwithpage{Parsing time
relative to shuffled} correspond to peaks in log file size in
fig.~\refwithpage{Parsing time, file size, and number of lines}, and
peaks in fig.~\refwithpage{parsing time vs file size vs number lines
factor}, where a peak means that performance is better.  The explanation
for this took some time to arrive at, but it turns out to be reasonably
simple.  The large log files in question were caused by a mail forwarding
loop, where the distribution of log lines is quite different to normal,
resulting in different performance characteristics.  The vast majority of
log lines when a mail loop occurs are from Postfix components which have a
small number of rules associated with them, whereas in general the majority
of log lines are from the Postfix component (\daemon{smtpd}) with the
highest number of rules.  Furthermore in normal log files the majority of
log lines are distributed much more evenly across rules than the majority
of log lines during a mail loop.

These two characteristics combine to reduce the average number of rules
required to parse a log line when there is a mail loop, as shown by the
peaks in fig.~\refwithpage{parsing time vs file size vs number lines
factor} and troughs in fig.~\refwithpage{Parsing time relative to
shuffled}.  When the rule ordering is reversed the majority of log lines
generated by a mail loop will be parsed using very few rules, because there
are very few rules for the Postfix components producing the log lines.
Without a mail loop the Postfix components producing the majority of log
lines have a large number of rules associated with them, and reverse
ordering deliberately starts with the least likely rules and moves towards
the most likely.  This leads to a noticeable drop in the average time
required to parse a log line during a mail loop, as shown in
fig.~\refwithpage{Parsing time relative to shuffled}.  The number of rules
which need to be consulted when the ordering is shuffled varies between the
optimal and the pessimal, with the performance varying proportionally.
